Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2018-12-06 scagnostics.Rmd
Editing sec 3.4.

- remind reader of the options for optimisation in guided tour
- how you use them here
-

The guided tour combines optimisation with interpolation between pairs of planes. Target planes of the path are chosen to maximise the PPI. There are three derivative-free optimisation methods available in the guided tour: \verb# search_better_random# (1), \verb# search_better# (2), and \verb# search_geodesic# (3). Method 1 casts a wide net randomly generating projection planes, computing the PPIs and keeping the best projection. Method 2 conducts a localised maximum search. Method 3 is quite different: a local search is conducted to determine a promising direction, and then this direction is followed until the maximum in that direction is found. In each case, a single target plane is chosen by the search methods, and a tour path is generated to interpolate from the current projection to the target plane.

@@

\textcolor{red}{Double challenging: fine structure + (coherent) noise}
\textcolor{red}{simple standard guided tour does not find pipe easily with holes index, did not try brute force approach used with TIC index yet}

As seen before (reference figure) MIC or TIC should be able to find the view down the pipe. The optimisation itself is however expected to be challenging because one has to be fairly close to the optimal projection to observe the increase in index value. We therefore suggest a two step procedure for the uncovering of such "small scale structures". The first step consists in "scouting" projections. Here we use the "search_better" optimisation available with the tourr package, with a standard search window $\alpha = 0.5$ and disabling the cooling (in practice this can be done by fixing the cooling parameter to $1$), together with a large number of tries (setting max.tries to 5000 here). The resulting search will quickly steer away from the most uninteresting projections, followed by a broad scan of the thus identified region. While the lack of cooling means that this will not converge towards the maximum, each step in this first search will increase the index value, thus it can be used to provide an informed starting plane that can be used in a second step where we use the "search_geodesic" optimization to improve the index further towards the maximum.
The described procedure is CPU time intensive, however it is promising that we manage to uncover such small scale structure in the distribution. On the other hand no "guarantee" can be provided for uncovering the special view. Parameters for the scouting phase will need to be adapted to different examples, and with increasing number of dimensions it may be better to consider several runs with different starting projections.
This gives some indication on how to handle difficult datasets. As we will see below most of the time standard "search_geodesic" optimisation will be efficient enough to uncover the interesting views.

@@

The results are shown below, first Fig. \ref{fig:pipeFirstRun} shows how the considered index values evolve in the scouting phase. Note that this is the interpolated path and there are small local maxima and minima along the index value, especially because we keep the step size in the guided tour large. We can see the clear maximum of the TIC index in the final projection, as well as clear minimum of the convex index that carries similar information On the other hand we see that MIC or the dcor2d index are not suitable for uncovering the special view here.

\textcolor{red}{Sometimes index value decreases, because index is not rotationally invariant.}
\textcolor{red}{Updated the plot, makes it look like skinny, 1-convex should be better index to optimise, however in practice this is not the case, probably because TIC is less sensitive to fluctuations, making it more robust against getting stuck in local maxima.}

>>>>

The guided tour combines optimisation with interpolation between pairs of planes. Target planes of the path are chosen to maximise the PPI. There are three derivative-free optimisation methods available in the guided tour: \verb# search_better_random# (1), \verb# search_better# (2), and \verb# search_geodesic# (3). Method 1 casts a wide net randomly generating projection planes, computing the PPIs and keeping the best projection. Method 2 conducts a localised maximum search. Method 3 is quite different: a local search is conducted to determine a promising direction, and then this direction is followed until the maximum in that direction is found. For all methods the optimisation is iterative, and we record each step where a plane with increased index value has been found. The stopping condition is given as number of tries that did not yield an increased index value at the final step in the optimisation.
For method 1 and 2 two parameters control the optimisation: the search window $\alpha$, giving the maximum distance from the current plane in which projections are samples, and the cooling factor, giving the incremental decrease in search window size. Method 3 in principle also has two free parameters, which are however fixed in the current implementation. The first is the small step size used when evaluating the most promising direction, it is fixed to 0.01, and the second parameter being the window over which the line search is performed, fixet to $\pm \pi/4$ away from the current plane.
For distributions and indexes with smooth behaviour and good squinatbaility, method 3 is the most efficient method for optimisation. If these two criteria are not met the method may still be useful, but only given an informed starting point. In such cases we can follow a method similar to that proposed by Friedman, i.e. we break the optimisation in two distinct steps. A first step ("scouting") uses method 2 with large search window and no cooling as a way of stepping over fluctuations and local maxima and yielding an approximation of the global maximum (note here that this also requires large number of tries for the stopping condition in particular in high dimensions, since most randomly picked planes will not be interesting). The second step is then to use method 3 starting from the approximate maximum, which will return the global maximum.

@@

As seen in Figure \ref{fig:plannedtour} the pipe structure has a small squint angle, and indeed using only method 3 does not uncover it in the six dimensional dataset. We therefore adapt the two step procedure described above. For the scouting we use $\alpha = 0.5$ and stopping condition of maximum 5000 tries, and we optimise the TIC index.

@@

The results are shown below, first Fig. \ref{fig:pipeFirstRun} shows how the considered index values evolve in the scouting phase.
\textcolor{red}{Updated the plot, makes it look like skinny, 1-convex should be better index to optimise, however in practice this is not the case, probably because TIC is less sensitive to fluctuations especially also under rotation, making it more robust against getting stuck in local maxima. Maybe it will be better to leave scales free for this plot, since increase in TIC is very hard to see currently. Or just remove plot? not actually that intersting...}
